Distortion corrected kernel density estimate on Riemannian manifolds

### 2022-11-06 paper/paper.tex
wrote intro

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of \(\mathbb{R}^d\). Applications of kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on. However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than \(\mathbb{R}^d\), kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted. \textcite{Pelletier2005-vu} propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known. Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold. Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than \(\mathbb{R}^d\). Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function. This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][ for details]{Heidenreich2013-bl,Scott2015-vl}.
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][ for reviews]{Jones1992-ef,Cao1994-st,Jones1996-cb,Wand1994-xu}. The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}. In this paper, we focus on the multivariate kernel density estimation.

Note that a fixed bandwidth matrix \(\pmb{H}\) is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
The bandwidth is varied depending on either the location of the sample points {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]} or that of the estimated points {[}balloon estimator; \textcite{Terrell1992-ut}{]}. In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} {[}VKDE; Section 6.6 of \textcite{Scott2015-vl}{]} unless otherwise stated.
However, these kernel density estimators are based on random samples in the Euclidean space.

For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, \textcite{Pelletier2005-vu} generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points. The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature \autocite{Henry2009-ll}.
However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space. Therefore, the kernel density estimator from \textcite{Pelletier2005-vu} is not applicable because the geodesic distance and the volume density function are unknown.
This is when we introduce manifold learning to reduce the input data dimension. For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in \textcite{Cheng2021-dh}), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure in recovering the geometry of the manifold \autocite{Goldberg2008-co}. To tackle this problem and measure the distortion incurred in manifold learning,
\textcite{Perrault-Joncas2013-pq} propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
The Riemannian metric using the method of \textcite{Perrault-Joncas2013-pq} gives some idea of the distortion of an embedding. Mapping the points through a non-linear function ``stretches'' some regions of space and ``shrinks'' others. The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.

By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding. Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point. Then the DC-KDE is used to estimate the density of the manifold and distortions induced by manifold learning methods are fixed with the estimated geometric information. Our distortion-corrected estimator is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold. These density estimates are useful in many other areas, including classification, clustering and anomaly detection. Similar to \textcite{Cheng2021-dh}, the highest density region plots\autocite{Hyndman1996-lk} could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds. Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection. \autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D twin peaks structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space. Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.

>>>>

Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}. For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well known curse of dimensionality. However the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d<<m\). The problem of KDE on manifolds has been well studied see \textcite{Pelletier2005-vu} (FIND SOME MORE REFS) for details. However, in these setting it is assumed that the underlying manifold in known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\). Details of these algorithms are summarised (OTHER REF) in \textcite{Cheng2021-dh}. One could easily apply KDE to the output vectors \(\pmb{y}\). However this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is

\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M\]

where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a change of variables, this would imply the following density for \(\pmb{y}\)

\[Pr(\pmb{y}\in\psi(\mathcal{A}))=\int_{\psi(\mathcal{A})} f(\psi^{-1}(\pmb{y}))|g(\pmb{y})|^{-1/2}d\pmb{y}\]
where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(g(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\). It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|g(\pmb{y})|^{-1/2}\) which can be considered roughly analogous to the Jacobian determinant. Since \(\psi\) will typically be a non-linear function, the value of \(|g(\pmb{y})|^{-1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms, but accounts for the distortion induced by the algorithm. For this reason we refer to it as the distortion corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of \(|g(\pmb{y})|^{-1/2}\), for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is that of anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualisation, our approach allows a two or three dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead this is an artefact of the dimension reduction algorithm `squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data. Furthermore, our estimator can be used for manifolds with elements that are are a variety of objects, for instance rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to an extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm. \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space. \autoref{application} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage. Conclusions and discussions are presented in \autoref{conclusion}. Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.
